
---
title: "Understanding the *factors* associated with **Loans at Prosper Marketplace**"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
  pdf_document:
    highlight: zenburn
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
---

# Prosper Dataset: *Top Down*
In this exersize, we deal with a dataset from
[Prosper](https://en.wikipedia.org/wiki/Prosper_Marketplace)

ProsperLoans dataset contains peer to peer low-medium sized loan transactions. Current dataset has over .11 million transactions, consisting schemas related to buyers, seller & the loan.
To understand the sructure of the data and criticality of each variable we do an Exploratory data analysis (EDA) on the structure of the data."

> EDA of the above, shall benefit us in a number of ways:

1. Tracking collinearty & multi-collinearity

2. Tracking hidden relationships

3. Checking for outliers & imputation of missing values 

We will also share the graphical represntation of some of the factors discussed. This is because, Data visualization is a pivotal part of data analysis. The below link explains the need:
[Anscombe Quartlet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet) 

************

#  Analyzing *schemas* of the dataset

Number of factors in the dataset
```{r,  echo  =  FALSE,  tidy  =  TRUE}
#Accepting Loan Analysis data
mydata  <-  read.csv("ProsperLoanData.csv",  header  =  TRUE)
ncol(mydata)
```


```{r,  echo  =  FALSE}
library(ggplot2)
# i have used the below url for this
# "http://rstudio-pubs-static.s3.amazonaws.com/3364_d1a578f521174152b46b19d0c83cbe7e.html"
#For table headings
brown.bold.italic.text  <-  element_text(face  =  "bold.italic"
                                         ,  color  =  "brown")
```


Some important variables around which our EDA revolves:

1.  Credit scores

2.  Delinquency Accounts

3.  Credit Accounts

4.  Borrower's income

5.  Investors

6.  Estimated Lossed

7.  Estimated Fees

8.  Lender Yields



*************

##   Does More friends mean More Investments from them?
```
We take 2 variables here:-

1.  InvestmentFromFriendsAmount
2.  InvestmentFromFriendsCount

with the intuition that more friends will bring more investments.
```
```{r,  echo  =  FALSE}
#Taking the variables "InvestmentFromFriendsAmount" 
# & "InvestmentFromFriendsCount"
friends  <-  mydata[,c("InvestmentFromFriendsAmount","InvestmentFromFriendsCount")]
#analyzing the variables- we get Large Tails in the right
i1  <-  fivenum(mydata$InvestmentFromFriendsAmount)
i2  <-  fivenum(mydata$InvestmentFromFriendsCount)
friends  <- na.omit(friends)
#Picking subscribers >0 friends & Investment
friends  <- subset(friends,   (InvestmentFromFriendsAmount  >  0  
                               &  InvestmentFromFriendsCount  >  0))
#Splitting the variables
investment  <-  as.vector(friends$InvestmentFromFriendsAmount)
friends.count  <-  as.vector(friends$InvestmentFromFriendsCount)
cor.friends  <-  cor(investment,  friends.count)
#checking z-value for Pearson's correlation to test its validity at 95% CI
cor.1  <-  cor.test(investment,  friends.count,  method  =  "kendall",  conf.level  =  .95)
#Plotting the finding
p1  <-  ggplot(friends,  aes(y  =  investment,  x  =  friends.count,
                             colour   =   friends.count))
p1  <-  p1  +  geom_line(size  =  1)  +  scale_colour_gradient(low  =  "red")
p1  <-  p1  +  xlab("Number of friends")  +  ylab("Investment by friends")
p1  <-  p1  +  ggtitle("Yes! \n More friends means More money") + 
     theme(plot.title  =  element_text(lineheight  =  1.8, face  =  "bold"))
p1  <-  p1  +  theme(panel.background  =  element_rect(fill  =  "lightblue"))
p1  <-  p1  +  stat_smooth(method  =  "glm",   fill  =  "blue"
                ,  colour  =  "darkblue",  size  =  2,  alpha  =  0.25)
p1  <-  p1  +  theme(aspect.ratio  =  1)
p1  <-  p1  +  theme(title = brown.bold.italic.text, axis.title = brown.bold.italic.text)
p1  <-  p1  +  annotate("text",  x  =  10,  y  =  15000,  color  =  "Dark Green"
               ,  label  =  "Smoothed line (Dark Blue)",  size  =  3)
p1
```

*************

##   Does Debt/Income correlates positively with Borrower rate?
```
We take 2 variables here:-

1.  BorrowerRate
2.  DebtToIncomeRatio

with the notion that borrowers with higher levels of existent debt relative 
to their income, shall get new loans at higher costs.
```

```{r,  echo  =  FALSE}
#Taking the variables "BorrowerRate" & "DebtToIncomeRatio"
debt  <-  mydata[,c("BorrowerRate","DebtToIncomeRatio")]
#Getting rid of NA entries to support convergence
debt  <-  na.omit(debt)
#Picking entries >0 bebt.rate & Debt/Income ratio
debt  <-  subset(debt,   (debt$BorrowerRate  >  0  
                               &  debt$DebtToIncomeRatio  >  0))
#Checking the spread of both variables
i3  <-  fivenum(debt$BorrowerRate, na.rm  =  TRUE)
i3  <-  fivenum(debt$DebtToIncomeRatio, na.rm  =  TRUE)
debt.to.income  <-  as.vector(debt$DebtToIncomeRatio)
debt.rate  <-  as.vector(debt$BorrowerRate)
#checking z-value for Pearson's correlation to test its validity at 95% CI
cor.debt  <-  cor(debt.to.income,  debt.rate)
```
Correlation between Debt/Income and Borrower rate is

0.06286032

Nothing significant from this data could be concluded
about the dependence of (Debt/Income) ratio on Borrower Rate



*************


##  Estimated Yield must vary proportionately with Estimated-Loss
```{r,  echo  =  FALSE}
#Taking the variables "EstimatedEffectiveYield" & "EstimatedLoss"
loss.yield  <-  mydata[,c("EstimatedLoss","EstimatedEffectiveYield")]
#Getting rid of NA entries to support convergence
loss.yield  <-  na.omit(loss.yield)
#Picking entries >0 bebt.rate & Debt/Income ratio
loss.yield  <-  subset(loss.yield,   (EstimatedEffectiveYield  >  0  
                               &  EstimatedLoss  >  0))
estimated.loss  <-  as.numeric(loss.yield$EstimatedLoss)
est.yield  <-  as.numeric(loss.yield$EstimatedEffectiveYield)
i5  <-  fivenum(estimated.loss)
i6  <-  fivenum(est.yield)
#Checking for Correlation between Expected Loss & Estimated yield
yield.cor  <-  cor(estimated.loss,  est.yield)
#Plotting the finding
p2  <-  ggplot(loss.yield,  aes(y  =  estimated.loss,  x  =  est.yield,
                             colour   =   estimated.loss))
p2  <-  p2  +  geom_line(size  =  1)  +  scale_colour_gradient(low  =  "red")
p2  <-  p2  +  xlab("Estimated yield")  +  ylab("Estimated Loss")
p2  <-  p2  +  ggtitle("Indeed! \n Estimated Loss increases Estimated Yield") + 
     theme(plot.title  =  element_text(lineheight  =  1.8,  face  =  "bold"))
p2  <-  p2  +  theme(panel.background  =  element_rect(fill  =  "lightblue"))
p2  <-  p2  +  stat_smooth(method  =  "gam",   fill  =  "blue"
                ,  colour  =  "darkblue",  size  =  2,  alpha  =  0.25)
p2  <-  p2  +  theme(aspect.ratio  =  1)
p2  <-  p2  +  theme(title  =  brown.bold.italic.text
                     ,  axis.title  =  brown.bold.italic.text)
p2  <-  p2  +  annotate("text",  x  =  .2,  y  =  .2,  color  =  "Dark Green"
               ,  label  =  "Smoothed line (Dark Blue)",  size  =  3)
p2
```



*************
##  Higher Prosper Score brings down Lender's Yield
```
We take the variable 

1.  ProsperScore

2.  LenderYield

To track how lender's yield gets altered by Prosper's score 
```
```{r,  echo  =  FALSE}
#Taking the variables "ProsperScore" & "LenderYield"
score.yield  <-  mydata[,c("ProsperScore","LenderYield")]
#Getting rid of NA entries to support convergence
score.yield  <-  na.omit(score.yield)
#Picking entries >0 bebt.rate & Debt/Income ratio
score.yield  <-  subset(score.yield,   (ProsperScore  >  0  
                               &  LenderYield  >  0))
prosper  <-  as.numeric(score.yield$ProsperScore)
yield  <-  as.numeric(score.yield$LenderYield)
i7  <-  fivenum(prosper)
i8  <-  fivenum(yield)
#doing a significance test to come out with the estimates
prosper.cor  <-  cor(prosper,  yield)
```


There is indeed a strong negative correlation
of 0.6497835
between Prosper Score & Lender's yield


*************
##  Amount Delinquent is less for people with high Income?
```
Deliquency is state of defaulting on a loan payment by greater than 30 days. This reduces the financial credibility of a borrower

Hence the intuition is that High income individuals 
should not be facing a hgh deliquency amount
```

```{r,  echo  =  FALSE}
#Taking the variables "ProsperScore" & "StatedMonthlyIncome"
delinquent.income  <-  mydata[,c("AmountDelinquent","StatedMonthlyIncome")]
#Getting rid of NA entries to support convergence
delinquent.income  <-  na.omit(delinquent.income)
#Picking entries >0 bebt.rate & Debt/Income ratio
delinquent.income  <-  subset(delinquent.income,   (AmountDelinquent  >  0  
                               &  StatedMonthlyIncome  >  0))
income  <-  as.numeric(delinquent.income$AmountDelinquent)
delinquent  <-  as.numeric(delinquent.income$StatedMonthlyIncome)
#span of the variables
i9  <-  fivenum(income)#Right skewed
i10  <-  fivenum(delinquent)#Outliers both ways
#Transforming the variables
income  <-  ifelse(income  >  100,  1,  0)
delinquent  <-  ifelse(delinquent  >  1000,  1,  0)
#doing a significance test to come out with the estimates
cor.delinquent  <-  cor(delinquent,  income)#low correlation
```


Surpringly, There is no strong correlation
between monthly income & Delinquent Amount
0.001861629


*************

##  Do borrower's with high ratings have lower number of Current Delinquencies

```
High credit ratings go hand in hand with borrowers who have been consistent with their bill payments.

Intuition: Individuals with high ratings should not be having a higher number deliquent accounts.
```

```{r,  echo  =  FALSE}
#Taking the variables "ProsperRating..numeric." & "CurrentDelinquencies"
rating.deliquencies  <-  mydata[,c("ProsperRating..numeric.",  
                              "CurrentDelinquencies")]
#Getting rid of NA entries to support convergence
rating.deliquencies  <-  na.omit(rating.deliquencies)
#Picking entries >0 Current Deliquencies
rating.deliquencies  <-  subset(rating.deliquencies,  CurrentDelinquencies  >  0)
rating  <-  as.numeric(rating.deliquencies$ProsperRating..numeric.)
delinquencies  <-  as.numeric(rating.deliquencies$CurrentDelinquencies)
#doing a significance test to come out with the estimates
ratings.cor  <-  cor.test(rating,  delinquencies,  conf.level  =  .95)#p<<.05 
#Plotting the finding
p3  <-  ggplot(rating.deliquencies,  aes(y  =  delinquencies,  x  =  rating,
                             colour   =   -delinquencies))
p3  <-  p3  +  geom_line(size  =  1)
p3  <-  p3  +  xlab("Prosper Rating")  +  ylab("Delinquencies")
p3  <-  p3  +  xlim(0,  8)   +  ylim(0,  60)  
p3  <-  p3  +  ggtitle("Rating Vs Number of Deliquencies") + 
     theme(plot.title  =  element_text(lineheight  =  1.8,  face  =  "bold"))
p3  <-  p3  +  theme(panel.background  =  element_rect(fill  =  "light yellow"))
p3  <-  p3  +  stat_smooth(method  =  "glm",   fill  =  "blue"
                ,  colour  =  "Dark Green",  size  =  2,  alpha  =  0.25)
p3  <-  p3  +  theme(aspect.ratio  =  1)
p3  <-  p3  +  theme(title  =  brown.bold.italic.text
                     ,  axis.title  =  brown.bold.italic.text)
p3  <-  p3  +  annotate("text",  x  =  7,  y  =  30,  color  =  "Dark Blue"
               ,  label  =  "Smoothed line (Dark Green)",  size  =  3)
p3
```


A significant relationship between Prosper Ratings and number of Deliquent accounts can't be established from the given data!


*************
##  Are multiple recommendations always given to people with higher income?
```{r,  echo  =  FALSE}
#Taking the variables "StatedMonthlyIncome" & "Recommendations"
income.reco  <-  mydata[,c("StatedMonthlyIncome",  "Recommendations")]
#Getting rid of NA entries to support convergence
income.reco  <-  na.omit(income.reco)
#Picking entries > 0 Recommendations
income.reco  <-  subset(income.reco,  Recommendations  >  0)
income  <-  as.numeric(income.reco$StatedMonthlyIncome)
recommendation.type  <-  as.numeric(income.reco$Recommendations)
#Transforming Reco into a binary variable <2, >=2 because of distribution
recommendation.type  <- ifelse(recommendation.type  <  2,  "Single",  "Multiple")
#preparing Box plots for the same
p4  <-  ggplot(income.reco,  aes(x  =  recommendation.type,  y  =  income,  
                                 fill  =  recommendation.type))
p4  <-  p4  +  geom_boxplot(aes(fill  =  factor(recommendation.type)))
p4  <-  p4  +  xlab("Recommendations")  +  ylab("income band") 
p4  <-  p4  +  ggtitle("No! \n Prosper with 1 recommendation \n earn more")  +
       theme(plot.title  =  element_text(lineheight  =  .8,  face  =  "bold"))
p4  <-  p4  +  theme(panel.background  =  element_rect(fill  =  "light blue"))
p4  <-  p4  +  theme(aspect.ratio  =  1.0)
p4  <-  p4  +  geom_jitter(aes(color = recommendation.type))
p4  <-  p4  +  theme(title = brown.bold.italic.text, axis.title = brown.bold.italic.text)
p4
```




*************
##  Do borrower's with higher Credit scores have lower APR?
```
we take the below variables, and study their correlation across all values 

1.  CreditScoreRangeUpper

2.  BorrowerAPR
```

```{r,  echo  =  FALSE}
#Taking the variables "CreditScoreRangeUpper" & "BorrowerAPR"
credit.APR  <-  mydata[,c("CreditScoreRangeUpper","BorrowerAPR")]
#Getting rid of NA entries to support convergence
credit.APR  <-  na.omit(credit.APR)
#Picking entries > 0 Recommendations
credit.APR  <-  subset(credit.APR,  BorrowerAPR  >  0)
credit  <-  as.numeric(credit.APR$CreditScoreRangeUpper)
apr  <-  as.numeric(credit.APR$BorrowerAPR)
#preparing Box plots for the same
#doing a significance test to come out with the estimates
cor.credit  <-  cor(credit,  apr)
apr.cor  <-  cor.test(credit,  apr,  conf.level = .95)#Highly significant
```

We have obtained a correlation of 0.4297073
, which indicates a strong negative correlation.

This brings clarity to the idea that higher credit scores get lower APR for the borrowers 


************

```{r,  echo  =  FALSE}
#detaching the needful packages
detach(package:ggplot2, unload  =  TRUE)
```


